The Bot on Speaking Terms: The Effects of Conversation Architecture on Perceptions of Conversational Agents

### 2023-02-23 figures/fig-heatmap_coverage.tex
update introduction & related work; descriptions for figures

\begin{figure*}[]
  \centering
  \includegraphics[width=\textwidth]{figures/fig-heatmap_coverage.pdf}
  \caption{Heatmap of literature for perception of conversational agents for each type of conversational architecture element}
  \label{fig:heatmap_coverage}
\end{figure*}

>>>>

\begin{figure*}[]
  \centering
  \includegraphics[width=\textwidth]{figures/fig-heatmap_coverage.pdf}
  \caption{Comparison between identified relationships and explored connections in literature}
  \Description{This figure shows a side by side comparison between the 183 identified relationships and the 265 explored connections in literature for conversation architecture elements and perceptions of agents. The visuals are shown as two heatmaps, with darker shades depicting higher number of relationships. The numbers within each cell at the interaction of a specific conversation architecture element and perception aspect outlines the number of relationships found in the reviewed corpus.}
  \label{fig:heatmap_coverage}
\end{figure*}

### 2023-02-23 figures/fig-heatmap_identified.tex
restructured manuscript
update introduction & related work; descriptions for figures

%% Side by side comparison between coverage and identified relationships

\begin{figure*}[]
  \centering
  \includegraphics[width=\textwidth]{figures/fig-heatmap_coverage.pdf}
  \caption{Heatmap of literature for perception of conversational agents for each type of conversational architecture element}
  \label{fig:heatmap_coverage}
\end{figure*}

>>>>

%% Heatmap of the identified relationships between perceptions & conversation architecture

\begin{figure*}[]
  \centering
  \includegraphics[width=\textwidth]{figures/fig-heatmap_identified.pdf}
  \caption{Synthesized framework on the identified relationship between perception of agents and conversation architecture elements}
  \Description{This figure shows a heatmap the 183 identified relationships between conversation architecture elements and perceptions of agents. Each speech element is listed down the rows of the heatmap, grouped by the four categories dialog strategy, content affectiveness, content style, and speech format. Each perception aspect is listed across the columns of the table, grouped by the four categories perception of interaction with agent, perception of agent's ability, perception of sociability with agent, and perception of agent's humanness. Each cell has a different shade of colour, with the darker shades depicting higher number of relationships, and the lighter shades depicting smaller number of relationships.}
  \label{fig:heatmap_identified}
\end{figure*}

### 2023-02-23 figures/fig-prisma.tex
update introduction & related work; descriptions for figures

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{figures/fig-prisma.pdf}
  \caption{Literature search using PRISMA guidelines}
  \label{fig:prisma}
\end{figure}

>>>>

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{figures/fig-prisma.pdf}
  \caption{Literature search process}
  \Description{This figure shows the literature search process, starting with automated databases search yielding 2901 publications. Screening the titles and abstracts of the papers based selection criteria resulted in 221 papers. Next, full article reviews filtered the list down to 49 papers. An additional search identified 8 more papers, for a total of 57 papers included in our reviewed corpus.}
  \label{fig:prisma}
\end{figure}

### 2023-02-23 sections/1-introduction.tex
update introduction & related work; descriptions for figures

Conversational agents (CAs), such as Amazon Alexa or Google Assistant, are designed to interact with humans using natural language through voice or text modalities. Since these agents communicate in natural language, \textit{conversation architecture}---what and how a CA says---affects users' perceptions of agents and, in turn, the effectiveness of their interactions~\cite{knijnenburg2016inferring, seeger2021chatbots}. According to previous research, there is a correlation between CAs' conversation architecture and users' perception of these agents.

While various studies investigated how different conversation architecture elements affect agents' perceptions, there is a paucity of knowledge on the overall coverage and effects of these relationships explored in literature. For conversation architecture elements, the use of affective language is commonly explored, with studies finding that conversational agents are perceived as more socially present and emotionally intelligent if they use sentiment-adaptive response based on user's utterances~\cite{diederich2019emulating, yang2017perceived}. Another commonly explored element is the use of prosody in voice-based agents, such as the use of express prosody contributing to higher perceived enjoyment and intimacy with the agent \cite{kim2020can}. However, there is a lack of research to synthesize these findings into an overall framework. One of the possible issues is the lack of clear and consistent definitions of the aspects of users' perceptions of agent and conversation architectural elements.

Currently, there is no systematic understanding on the elements of conversation architecture that are relevant to the perceptions of agents. Some papers have started to examine the speech specifics explored in literature, such as \citet{clark2019state} capturing different elements of system speech production for voice-based agents and categorizing them based on content, style, and other topics. There are also some investigations of conversation architecture elements within the broader field of studying cues that trigger a social reaction of the user towards the CA. Feine et al's \cite{feine2019taxonomy} paper laid the foundation to create a consistent taxonomy for these social cues. However, there is no comprehensive research specifically for conversation architecture elements that are relevant to user perceptions.

Related to user perceptions, there is also no systematic understanding on the perceptions of agents affected by speech variations. Some research discussed the perceptions of agents as part of the broader scope of user experiences of CAs. These publications have gathered UX dimensions used to evaluate conversational agents, which include the perceptions of agents (e.g. empathy, humanness) alongside other assessments (e.g. physiological data) \cite{clark2019state, finch2020towards}. Some even categorized the metrics based on the type of method used, including technical measures like system error rate and perception measures like perceived social presence \cite{zheng2022ux}. However, these studies are assessing the general field of UX measurements and not specifically on the perceptions of agents, and do not analyze the effects of conversation architecture on these perceptions.

These research gaps discussed above: the paucity of knowledge on synthesizing the effect of speech specifics on perceptions of CAs as well as the lack of systematic understanding on the elements of conversation architecture and perceptions of agents motivated us to the goal of how to design conversation architecture in order to orchestrate perceptions of conversational agents. First, we want to understand \textbf{(RQ1) what perception aspects of agents are being explored in relation to the effects of conversation architecture elements}. While there are various architecture elements of conversation architecture used in a CA's utterances, we don't have a holistic understanding of them. This leads to our second research question: \textbf{(RQ2) what elements of conversation architecture are relevant to the perceptions of agents?} Lastly, we want to understand the effect of conversation architecture on perceptions of agent, specifically: \textbf{(RQ3) what are the relationships between conversation architecture elements and the perceptions of CAs?}

To answer these questions, we performed a comprehensive review on existing literature from 2010 to 2022 and synthesized the effect of conversation architecture elements on the perceptions of CAs based on 57 relevant papers (\autoref{fig:heatmap_identified}). We also grouped the perceptions of agents affected by conversation architecture into four categories and eleven aspects (\autoref{tbl:perceptions}), and grouped the speech variations related to perceptions of agents into four categories and eleven elements (\autoref{tbl:conversation_architecture}). %Lastly, we coded the explored connections between conversation architecture and perceptions and synthesized their relationships into a framework. Through our analysis, we also found that there are inconsistencies in the assessment of perceptions across studies, as well as influencing factors other than conversation architecture that impact the perceptions of agents. 
This work contributes to the HCI community by presenting a framework that demonstrates the effect of conversation architecture on perceptions of conversational agents, as well as taxonomies for perception aspects and conversation architecture elements providing clear and consistent terms that can be used in future research. Together, these contributions move us one step closer to designing CAs that effectively interact with users. 
%Also, our analysis extends existing reviews on UX research of CAs \cite{clark2019state}\cite{rapp2021human}\cite{zheng2022ux} by providing a synthesized framework outlining the relationships between conversation architecture and perceptions of agents. With the help of this framework, we are taking the first step to understand how to design conversational agents to solicit specific social reactions from users.

>>>>

Conversational agents (CAs), such as Amazon Alexa or Google Assistant, are designed to interact with humans using natural language through voice or text modalities. Since these agents communicate in natural language, \textit{conversation architecture}---what and how a CA says---affects users' perceptions of agents and, in turn, the effectiveness of their interactions in achieving users' goals \cite{knijnenburg2016inferring, seeger2021chatbots}. According to previous research, there is a correlation between CAs' conversation architecture and users' perception of these agents.

While various studies investigated how different speech elements affect agents' perceptions, there is a paucity of knowledge on the overall coverage and effects of these relationships explored in literature. One reason for this disparity is the lack of clear and consistent definitions of the aspects of users' perceptions of agent and conversation architectural elements. Some papers have discussed the speech specifics of agents as part of  a broader analysis on the state of research \cite{clark2019state} or social cues \cite{feine2019taxonomy}, but there is no comprehensive research specifically for conversation architecture elements that are relevant to user perceptions. Similarly for user perceptions, while there some research exploring the perceptions of agents as part of understanding the user experiences \cite{clark2019state, finch2020towards} or social cues \cite{feine2019taxonomy} of CAs, there is no exhaustive research specifically for perceptions of agents that are affected by conversation architecture elements.

%While various studies investigated how different conversation architecture elements affect agents' perceptions, there is a paucity of knowledge on the overall coverage and effects of these relationships explored in literature. For conversation architecture elements, the use of affective language is commonly explored, with studies finding that conversational agents are perceived as more socially present and emotionally intelligent if they use sentiment-adaptive response based on user's utterances~\cite{diederich2019emulating, yang2017perceived}. Another commonly explored element is the use of prosody in voice-based agents, such as the use of express prosody contributing to higher perceived enjoyment and intimacy with the agent \cite{kim2020can}. However, there is a lack of research to synthesize these findings into an overall framework. One of the possible issues is the lack of clear and consistent definitions of the aspects of users' perceptions of agent and conversation architectural elements.

%Currently, there is no systematic understanding on the elements of conversation architecture that are relevant to the perceptions of agents. Some papers have started to examine the speech specifics explored in literature, such as \citet{clark2019state} capturing different elements of system speech production for voice-based agents and categorizing them based on content, style, and other topics. There are also some investigations of conversation architecture elements within the broader field of studying cues that trigger a social reaction of the user towards the CA. Feine et al's \cite{feine2019taxonomy} paper laid the foundation to create a consistent taxonomy for these social cues. However, there is no comprehensive research specifically for conversation architecture elements that are relevant to user perceptions.

%Related to user perceptions, there is also no systematic understanding on the perceptions of agents affected by speech variations. Some research discussed the perceptions of agents as part of the broader scope of user experiences of CAs. These publications have gathered UX dimensions used to evaluate conversational agents, which include the perceptions of agents (e.g. empathy, humanness) alongside other assessments (e.g. physiological data) \cite{clark2019state, finch2020towards}. Some even categorized the metrics based on the type of method used, including technical measures like system error rate and perception measures like perceived social presence \cite{zheng2022ux}. However, these studies are assessing the general field of UX measurements and not specifically on the perceptions of agents, and do not analyze the effects of conversation architecture on these perceptions.

These research gaps related to the paucity of knowledge on synthesizing the effect of speech specifics on perceptions of CAs as well as the lack of systematic understanding on the elements of conversation architecture and perceptions of agents motivated us to the goal of how to design conversation architecture in order to orchestrate perceptions of conversational agents. First, we want to understand \textbf{(RQ1) what perception aspects of agents are being explored in relation to the effects of conversation architecture elements}. While there are various architecture elements of conversation architecture used in a CA's utterances, we don't have a holistic understanding of them. This leads to our second research question: \textbf{(RQ2) what elements of conversation architecture are relevant to the perceptions of agents?} Lastly, we want to understand the effect of conversation architecture on perceptions of agent, specifically: \textbf{(RQ3) what are the relationships between conversation architecture elements and the perceptions of CAs?}

To answer these questions, we performed a comprehensive review on existing literature from 2010 to 2022 and synthesized the effect of conversation architecture elements on the perceptions of CAs based on 57 relevant papers (\autoref{fig:heatmap_identified}). We also grouped the perceptions of agents affected by conversation architecture into four categories and eleven aspects (\autoref{tbl:perceptions}), and grouped the speech variations related to perceptions of agents into four categories and eleven elements (\autoref{tbl:conversation_architecture}). 
We found that perceptions of interaction with CAs (usability, engagement, satisfaction) have the most number of explored and identified relationships with speech variations, while the perceptions of CAs' ability (intelligence, competence, credibility) have the least number of explored and identified relationships. Also, the conversation architecture elements of response delay and agent initiated dialog related perceptions of CAs are under-explored in literature.
%Lastly, we coded the explored connections between conversation architecture and perceptions and synthesized their relationships into a framework. Through our analysis, we also found that there are inconsistencies in the assessment of perceptions across studies, as well as influencing factors other than conversation architecture that impact the perceptions of agents. 
This work contributes to the HCI community by presenting a framework that demonstrates the effect of conversation architecture on perceptions of conversational agents, as well as taxonomies for perception aspects and conversation architecture elements providing clear and consistent terms that can be used in future research. Together, these contributions move us one step closer to designing CAs that effectively interact with users. 
%Also, our analysis extends existing reviews on UX research of CAs \cite{clark2019state}\cite{rapp2021human}\cite{zheng2022ux} by providing a synthesized framework outlining the relationships between conversation architecture and perceptions of agents. With the help of this framework, we are taking the first step to understand how to design conversational agents to solicit specific social reactions from users.

### 2023-02-23 sections/2-related_work.tex
update introduction & related work; descriptions for figures

There have been numerous studies on the user experiences for conversational agents, and some attempts to synthesize findings across literature. Several papers have mapped out the current state of research for conversational agents for specific agent types (e.g. text-based, voice-based, polyadic) highlighting major trends, topics, methods and evaluating metrics \cite{clark2019state, rapp2021human, zheng2022ux}\cmt{[33][5]}. Other papers looked at specific dimensions of user experiences or domains of usage, such as Van Pinxteren et al. \cite{van2020human} examining the effects of agents' human-like communicative behaviours on their relationships with users, and Kocaballi et al. \cite{kocaballi2022design} exploring the challenges and opportunities of CAs in healthcare. There are also publications aimed to unify concepts used in the research for CA user experiences, for instance Feine et al. \cite{feine2019taxonomy} creating a taxonomy of social cues, and Finch et al. \cite{finch2020towards} analyzing evaluation protocols for dialogue systems. To the best of our knowledge, there is currently no synthesis on the speech specifics of CAs, the perceptions of agents, or the relationship between them.

@@

>>>>

There are a number of studies investigating factors affecting user experiences for conversational agents, such as the relationship between conversation architecture elements and agents' perceptions. For instance, the use of affective language is commonly explored, with studies finding that conversational agents are perceived as more socially present and emotionally intelligent if they use sentiment-adaptive response based on user's utterances \cite{diederich2019emulating, yang2017perceived}. Another commonly explored element is the use of prosody in voice-based agents, such as the use of express prosody contributing to higher perceived enjoyment and intimacy with the agent \cite{kim2020can}. However, there is a paucity of research to synthesize these findings on user experiences with agents across literature.

In recent years, several papers have synthesized the current state of research for conversational agents for specific agent types (e.g. text-based, voice-based, polyadic) highlighting major trends, topics, methods and evaluating metrics \cite{clark2019state, rapp2021human, zheng2022ux}\cmt{[33][5]}. Other papers looked at specific dimensions of user experiences or domains of usage, such as \citet{van2020human} examining the effects of agents' human-like communicative behaviours on their relationships with users, and \citet{kocaballi2022design} exploring the challenges and opportunities of CAs in healthcare. There are also publications aimed to unify concepts used in the research for CA user experiences, for instance \citet{feine2019taxonomy} creating a taxonomy of social cues, and \citet{finch2020towards} analyzing evaluation protocols for dialogue systems. To the best of our knowledge, there is currently no synthesis on the speech specifics of CAs, the perceptions of agents, or the relationship between them.

@@

%Currently, there is no systematic understanding on the elements of conversation architecture that are relevant to the perceptions of agents. Some papers have started to examine the speech specifics explored in literature, such as \citet{clark2019state} capturing different elements of system speech production for voice-based agents and categorizing them based on content, style, and other topics. There are also some investigations of conversation architecture elements within the broader field of studying cues that trigger a social reaction of the user towards the CA. Feine et al's \cite{feine2019taxonomy} paper laid the foundation to create a consistent taxonomy for these social cues. However, there is no comprehensive research specifically for conversation architecture elements that are relevant to user perceptions.

%Related to user perceptions, there is also no systematic understanding on the perceptions of agents affected by speech variations. Some research discussed the perceptions of agents as part of the broader scope of user experiences of CAs. These publications have gathered UX dimensions used to evaluate conversational agents, which include the perceptions of agents (e.g. empathy, humanness) alongside other assessments (e.g. physiological data) \cite{clark2019state, finch2020towards}. Some even categorized the metrics based on the type of method used, including technical measures like system error rate and perception measures like perceived social presence \cite{zheng2022ux}. However, these studies are assessing the general field of UX measurements and not specifically on the perceptions of agents, and do not analyze the effects of conversation architecture on these perceptions.

### 2023-02-23 sections/3-method.tex
update introduction & related work; descriptions for figures

The initial query retrieved 2901 unique publications. We screened the titles and abstracts of the papers based on the selection criteria above, resulting in 221 papers. The fully body of these papers are then reviewed, selecting papers that met the criteria. A search for additional literature was also performed to supplement the corpus, adding 8 more papers to the selection. In total, we identified 57 relevant articles for analysis (see Figure \ref{fig:prisma}).

@@

The papers reviewed (n=57) were published between May 2011 and November 2022. More than 75\% of the papers were published in or after 2019, with a slight trend upwards over the years (see Figure \ref{fig:paper}). The vast majority of the papers in our corpus were published in conference proceedings (n=50), with the remainder papers published in journals. Out of the papers published in conferences, the top conferences were The ACM Conference on Human Factors in Computing Systems\footnote{https://dl.acm.org/conference/chi} (n=16), Conversational User Interfaces\footnote{https://dl.acm.org/conference/cui} (n=7), Human-Agent Interaction\footnote{https://dl.acm.org/conference/hai} (n=4), and Intelligent Virtual Agents\footnote{https://dl.acm.org/conference/iva} (n=4). 
%Out of the papers published in journals, some venues include the International Journal of Human-Computer Interaction\footnote{https://www.tandfonline.com/journals/hihc20} (n=2) and Interacting with Computers\footnote{https://academic.oup.com/iwc} (n=1).

>>>>

The initial query retrieved 2901 unique publications. We screened the titles and abstracts of the papers based on the selection criteria above, resulting in 221 papers. The fully body of these papers are then reviewed, selecting 49 papers that met the criteria. A search for additional literature was also performed to supplement the corpus, adding 8 more papers to the selection. In total, we identified 57 relevant articles for analysis (see \autoref{fig:prisma}).

@@

The papers reviewed (n=57) were published between May 2011 and November 2022. More than 75\% of the papers were published in or after 2019, with a slight trend upwards over the years (see \autoref{fig:paper}). The vast majority of the papers in our corpus were published in conference proceedings (n=50), with the remainder papers published in journals. Out of the papers published in conferences, the top conferences were The ACM Conference on Human Factors in Computing Systems\footnote{https://dl.acm.org/conference/chi} (n=16), Conversational User Interfaces\footnote{https://dl.acm.org/conference/cui} (n=7), Human-Agent Interaction\footnote{https://dl.acm.org/conference/hai} (n=4), and Intelligent Virtual Agents\footnote{https://dl.acm.org/conference/iva} (n=4). 
%Out of the papers published in journals, some venues include the International Journal of Human-Computer Interaction\footnote{https://www.tandfonline.com/journals/hihc20} (n=2) and Interacting with Computers\footnote{https://academic.oup.com/iwc} (n=1).

### 2023-02-23 sections/4-findings.tex
update introduction & related work; descriptions for figures

As shown in Table \ref{tbl:perceptions}, there are four categories defined for the perceptions of conversational agents covering 11 different aspects of perception. The details of each category and aspect are discussed below.

@@

As shown in Table \ref{tbl:conversation_architecture}, there are four categories defined for conversation architecture across 12 different elements. The details of each category and element are discussed below.

>>>>

As shown in \autoref{tbl:perceptions}, there are four categories defined for the perceptions of conversational agents covering 11 different aspects of perception. The details of each category and aspect are discussed below.

@@

As shown in \autoref{tbl:conversation_architecture}, there are four categories defined for conversation architecture across 12 different elements. The details of each category and element are discussed below.

### 2023-02-23 sections/5-discussion.tex
update introduction & related work; descriptions for figures

Our synthesized framework (Figure \ref{fig:heatmap-impact}) serves as a foundational understanding on the effect of conversation architecture on the perception of agents. It showcases the density of explored relationships between them, as well as highlighting areas that are under-explored. In order to continue building our knowledge, it is important to investigate further into the effect of contextual factors as well as multiple conversation architecture elements on users' perceptions.

>>>>

Our synthesized framework (\autoref{fig:heatmap_identified}) lays the foundation in understanding the effect of conversation architecture on the perception of agents. It showcases the density of explored relationships between them, as well as highlighting areas that are under-explored. In order to continue building our knowledge, it is important to investigate further into the effect of contextual factors as well as multiple conversation architecture elements on users' perceptions.

### 2023-02-23 tables/tbl-conversation_architecture.tex
update introduction & related work; descriptions for figures

& Text Formatting & capitalization, emoticons & 6
& \cite{fadhil2018effect}\cmt{[52]}\cite{kim2019comparing}\cmt{[89]}\cite{seeger2021chatbots}\cmt{[35]}\cite{volkel2022user}\cmt{[75]}\cite{westerman2019believe}\cmt{[9]}\cite{wilhelm2022keep}\cmt{[28]}
\\ \Xhline{1.2pt}
\end{tabular}%
}
\caption{Conversation architecture elements}
\label{tbl:conversation_architecture}
\end{table*}

>>>>

& Text Formatting & capitalization, emoticons & 6
& \cite{fadhil2018effect}\cmt{[52]}\cite{kim2019comparing}\cmt{[89]}\cite{seeger2021chatbots}\cmt{[35]}\cite{volkel2022user}\cmt{[75]}\cite{westerman2019believe}\cmt{[9]}\cite{wilhelm2022keep}\cmt{[28]}
\\ \Xhline{1.2pt}
\end{tabular}%
}
\caption{Conversation architecture elements}
\Description{This table categorizes conversation architecture elements into four categories,  provide some examples for the element, and show the number of papers as well as references to the papers in our reviewed corpus. The first category is dialog strategy, which contains the elements of social dialog, initiative and response delay. The second category is content affectiveness, which contains the elements of affective language and humour. The third category is content style, which contains the elements of formality, alignment and elaborateness. The fourth category is speech format, which contains the elements of prosody, disfluency, and text formatting.}
\label{tbl:conversation_architecture}
\end{table*}

### 2023-02-23 tables/tbl-perceptions.tex
update introduction & related work; descriptions for figures

    & Trust & credibility, trustworthy, truthfulness, confidence & 15

@@

>>>>

    & Credibility & trustworthy, truthfulness, confidence & 15

@@

\Description{This table categorizes perceptions of agents into four categories, provide some examples for the perception aspect, and show the number of papers as well as references to the papers in our reviewed corpus. The first category is perception of interaction with agent, which contains the aspects of usability, engagement, and satisfaction. The second category is perception of agent's ability, which contains the aspects of intelligence, competence, and credibility. The third category is perception of sociability with agent, which contains the aspects of conversation tone, social presence, and intimacy. The fourth category is perception of agent's humanness, which contains aspects of human-likeness and personality traits.}
